---
output: html_document
---

# Manipulating text {#text}

We introduced text in the previous chapter. In this chapter, we will show how to manipulate text as strings and factors. We will use the `states` dataset from the `poliscidata` package. For more on this data, see Chapter \@ref(datadownload).

```{r}
library("poliscidata")

states <- states
```

You can use `View(states)` to get a sense of the 50 observations and the 135 variables.

## Strings

We will use the package `string`. It is part of the `tidyverse` but can also be called individually. As you already should have installed the `tidyverse` by now, it is not necessary to install the package again.

```{r}
library("stringr")
```

Some of the functions have relatively simple purposes, such as `str_to_upper()` (which convert all characters in a string to upper case) and `str_to_lower()` (which convert all characters in a string to lower case).

```{r}
str_to_upper("Quantitative Politics with R")
```

```{r}
str_to_lower("Quantitative Politics with R")
```

Here, we are going to look at cigarette taxes, and namely on whether the cigarette taxes are in the low, middle or high category. To look at this we will use the `cig_tax12_3` variable in the `states` data frame.

```{r eval=FALSE}
table(states$cig_tax12_3)
```

We can see that the names for these categories are LoTax, MidTax and HiTax. With the code below we use `str_replace_all()` to replace the characters with new characters, e.g. HiTax becomes High taxes.

```{r}
states$cig_taxes <- str_replace_all(states$cig_tax12_3, 
                                    c("HiTax" = "High taxes", 
                                      "MidTax" = "Middle taxes",
                                      "LoTax" = "Low taxes"))

table(states$cig_taxes)
```

For examples on more of the functions available in the `stringr` package, see [this introduction](https://cran.r-project.org/web/packages/stringr/vignettes/stringr.html).

## Factors

For the cigarette taxes we have worked with above, these are categorical data that we can order. To work with ordered and unordered categories, factors is a class in `R` class that makes these categories good to work with. For factors, we are going to use the package `forcats`. This package is also part of the `tidyverse`.

```{r}
library("forcats")
```

We create a new variable, `cig_taxes_cat` as a factor variable and then we see what levels we have (and the order of these).

```{r}
states$cig_taxes_cat <- factor(states$cig_taxes)

levels(states$cig_taxes_cat)
```

As we can see, these levels are now in the wrong order (sorted alphabetically). We can use the `fct_relevel()` to specify the order of the categories (from low to high).

```{r}
states$cig_taxes_cat <- fct_relevel(states$cig_taxes_cat, 
                                    "Low taxes", 
                                    "Middle taxes", 
                                    "High taxes")

levels(states$cig_taxes_cat)
```

This will become useful later on when we want to make sure that the categories in a data visualisation has the correct order.

For additional guidance on the functions available in the `forcats` package, see [https://forcats.tidyverse.org/](https://forcats.tidyverse.org/).