---
output: html_document
---

# Manipulating text {#text}

We introduced text in the previous chapter. In this chapter, we will show how to manipulate text as strings and factors. We will use the `states` dataset from the `poliscidata` package. For more on this data, see Chapter \@ref(datadownload).

```{r}
library("poliscidata")

states <- states
```

You can use `View(states)` to get a sense of the 50 observations and the 135 variables.

## Strings

There are a few functions that are great to use, when you use strings. First, `paste()` makes it easy to connect two strings.

```{r}
paste("Hello", "World")
```

As you can see, there is a space between the two strings we that we connect. If you would like not to have a space between the two strings, we can use `paste0()`.

```{r}
paste0("face", "book")
```

Most of the relevant functions we can use will be in the package `stringr`. It is part of the `tidyverse` but can also be called individually. As you have already installed the `tidyverse` by now, it is not necessary to install the package again.

```{r}
library("stringr")
```


### Regular expression (regex)

When working with strings, and in particular when manipulating strings, it is useful to rely on regular expression. Regular expression is a formal language for specifying text strings. Before you can fully leverage the functions available in the `stringr` package, we suggest that you familiarize yourself with the basics of regular expression (regex for short).

What we will do here is to briefly introduce the basics. If you would like additional material, we recommend that you check out [https://github.com/ziishaned/learn-regex](https://github.com/ziishaned/learn-regex) and [https://github.com/aloisdg/awesome-regex](https://github.com/aloisdg/awesome-regex). 

To illustrate the usefulness of regex, we will use the functions `grep()` and `grepl()`. The former gives us information about the place of a string in a vector. The latter returns a logical vector with information on whether the text we are looking for is present or not. Let us save four pieces of text in an object called `trump_text`.

```{r}
trump_text <- c("Trump", "trump", "trump is a 0", "Trump is a loser")
```

To see where "trump" is mentioned in this object, we can first use `grep()`. Notice that we first specify the text we will like to find and then the object.

```{r}
grep("trump", trump_text)
```

The output shows that "trump" is present in the second and third place in the object. We can use `grepl()` to get a similar result.

```{r}
grepl("trump", trump_text)
```

The only difference here is that we get a `TRUE` or `FALSE` for each element in the object. 

Let us say, however, that we want to get information on whether "trump" or "Trump" is mentioned, i.e. that we do not care about whether it is with a lower t or not. To do this, we can use [] as a disjunction. Within this, we can say that we are looking for both "trump" and "Trump".

```{r}
grepl("[Tt]rump", trump_text)
```

The output not shows that either Trump or trump is present in all four elements. Next, we can use the same logic to explore whether a number is present in any of the elements.

```{r}
grepl("[0123456789]", trump_text)
```

We can see that a number is present in the third element. However, we do not need to include all numbers and we will get the same result by specifying a range of numbers from 0 to 9 (the same logic can be applied to letters, e.g. [a-o] will include all letters from a to o in the alphabet.

```{r}
grepl("[0-9]", trump_text)
```

If you want to check whether at least one of two different words is mentioned, e.g. Trump or trump, you can use the "|"-sign.

```{r}
grepl("Trump|trump", trump_text)
```

A few other characters that are good to know about:

* "." means any character. Tr.mp will match both Trump and Tramp.
* "\\." is a period (i.e., the "\" sign escapes the regex)
* "?" is for a character that might be there or not (for example, colou?r matches both color and colour)
* "\*" will be any number of similar characters (for example, wo*w will match wow, woow, wooow etc.)

### Changing the case of strings

Some of the functions have relatively simple purposes, such as `str_to_upper()` (which convert all characters in a string to upper case), `str_to_lower()` (which convert all characters in a string to lower case), `str_to_title()` (which convert the first letter in each word in a string to upper case) and `str_to_sentence()` (which convert the first letter in a sentence to upper case and everything else to lower case).

```{r}
str_to_upper("Quantitative Politics with R")
```

```{r}
str_to_lower("Quantitative Politics with R")
```

```{r}
str_to_title("Quantitative Politics with R")
```

```{r}
str_to_sentence("Quantitative Politics with R")
```

### Subset and mutate strings

We can use `str_sub()` to get a part of the text we are looking at. Say we want to get the first four characters of a string, we can specify `start = 1` and `end = 4`.

```{r}
str_sub("Quantitative Politics with R", start = 1, end = 4)
```

If we would like to get the last four characters, we can simply specify `start = -4` as the option.

```{r}
str_sub("Quantitative Politics with R", start = -4)
```

Here, we are going to look at cigarette taxes, and namely on whether the cigarette taxes are in the low, middle or high category. To look at this we will use the `cig_tax12_3` variable in the `states` data frame.

```{r eval=FALSE}
table(states$cig_tax12_3)
```

We can see that the names for these categories are LoTax, MidTax and HiTax. With the code below we use `str_replace_all()` to replace the characters with new characters, e.g. HiTax becomes High taxes.

```{r}
states$cig_taxes <- str_replace_all(states$cig_tax12_3, 
                                    c("HiTax" = "High taxes", 
                                      "MidTax" = "Middle taxes",
                                      "LoTax" = "Low taxes"))

table(states$cig_taxes)
```

For examples on more of the functions available in the `stringr` package, see [this introduction](https://cran.r-project.org/web/packages/stringr/vignettes/stringr.html).

Three other functions that can come in useful are from the `tidyr` package: `separate()`, `unite()` and `extract()`. `tidyr` is also part of the `tidyverse` and if you load the `tidyverse` package, you do not need to load `tidyr`. These functions come with multiple options and we suggest that you consult the documentations for these in order to see examples and the different options, e.g. `help("separate")`.

## Factors

For the cigarette taxes we have worked with above, these are categorical data that we can order. To work with ordered and unordered categories, factors is a class in `R` class that makes these categories good to work with. In brief, categorical data is a variable with a fixed set of possible values. This is also useful when you want to use a non-alphabetical ordering of the values in a variable.

For factors, we are going to use the package `forcats`. This package is also part of the `tidyverse`.

```{r}
library("forcats")
```

We create a new variable, `cig_taxes_cat` as a factor variable and then we see what levels we have (and the order of these).

```{r}
states$cig_taxes_cat <- factor(states$cig_taxes)

levels(states$cig_taxes_cat)
```

As we can see, these levels are now in the wrong order (sorted alphabetically). We can use the `fct_relevel()` to specify the order of the categories (from low to high).

```{r}
states$cig_taxes_cat <- fct_relevel(states$cig_taxes_cat, 
                                    "Low taxes", 
                                    "Middle taxes", 
                                    "High taxes")

levels(states$cig_taxes_cat)
```

This will become useful later on when we want to make sure that the categories in a data visualisation has the correct order.

For additional guidance on the functions available in the `forcats` package, see [https://forcats.tidyverse.org/](https://forcats.tidyverse.org/).

## Dates and time

To work with dates and time in `R`, there are two useful packages. The first is `hms` that is good with **h**ours, **m**inutes and **s**econds. The second is `lubridate` that is good with dates. Let us take a closer look at how to work with seconds, minutes and hours by loading the package `hms`.

```{r}
library("hms")
```

This package is useful if you want to easily convert minutes into hours, for example 500 minutes into hours. As `lubridate` also have an `hms()`, we will use `::` to tell that we are using the function `hms()` in the `hms` package.

```{r}
hms::hms(min = 500)
```

As we can see, this gives 8 hours and 20 minutes. We can also specify hours, minutes and seconds to get the time as POSIXct.

```{r}
hms::hms(hour = 15, min = 90, seconds = 12)
```

For dates, we will first load the package `lubridate`.

```{r}
library("lubridate")
```

This package has several functions that are useful in terms of working with dates. For example, we can use `ymd()` if we have text that has the year, month and day.

```{r}
ymd("2019/09/30")
```

The package can also work with months as text, such as:

```{r}
mdy("September 30, 2019")
```

The good thing about this is that we can work with the date information. Let us first save the date in an object called `date` and use `year()` to get the year out of the variable.

```{r}
# Save September 30, 2019 in object
date <- ymd("2019-09-30")

# Get year
year(date)
```

Similarly, we can get the week number out of the date.

```{r}
week(date)
```

We can see that this date was in week 39. We can use `wday()` to get the number of the day in the week this was.

```{r}
wday(date)
```

If we would rather prefer the name of the day, we can use `label` and `abbr` as options (the latter option in order to get the full day name).

```{r}
wday(date, label = TRUE, abbr = FALSE)
```

If you would like to get the difference between two dates, you can simply subtract one date from the other as in the example below.

```{r}
ymd("2019-09-30") - ymd("2019-09-01")
```

Some of the relevant functions in the `lubridate` package are: `year()` (year), `month()` (month), `week()` (week number), `day()` (day of month), `wday()` (day week), `qday()` (day of quarter), and `yday()` (day of year).

Last, if you have information on a date in a different format (i.e. not as used above), you can specify this format. For example, if the data is 30/09/2020, you can specify the format with %d/%m/%Y to turn it into a date.

```{r}
as.Date("30/09/2020", format = "%d/%m/%Y")
```

Here is a list of some of the most useful formats:

  - %d = day as number (0-31)
  - %a = abbreviated weekday
  - %A = unabbreviated weekday
  - %m = month (00-12)
  - %b = abbreviated month
  - %B = unabbrevidated month
  - %y = 2 digit year
  - %Y = four digit year