---
output: html_document
---

# Manipulating text {#text}

We introduced text in the previous chapter. In this chapter, we will show how to manipulate text as strings and factors. We will use the `states` dataset from the `poliscidata` package. For more on this data, see Chapter \@ref(datadownload).

```{r}
library("poliscidata")

states <- states
```

You can use `View(states)` to get a sense of the 50 observations and the 135 variables.

## Strings

We will use the package `string`. It is part of the `tidyverse` but can also be called individually. As you already should have installed the `tidyverse` by now, it is not necessary to install the package again.

```{r}
library("stringr")
```


Some of the functions have relatively simple purposes, such as `str_to_upper()` (which convert all characters in a string to upper case) and `str_to_lower()` (which convert all characters in a string to lower case).

```{r}
str_to_upper("Quantitative Politics with R")
```

```{r}
str_to_lower("Quantitative Politics with R")
```

We can use `str_sub()` to get a part of the text we are looking at. Say we want to get the first four characters of a string, we can specify `start = 1` and `end = 4`.

```{r}
str_sub("Quantitative Politics with R", start = 1, end = 4)
```

If we would like to get the last four characters, we can simply specify `start = -4` as the option.

```{r}
str_sub("Quantitative Politics with R", start = -4)
```

Here, we are going to look at cigarette taxes, and namely on whether the cigarette taxes are in the low, middle or high category. To look at this we will use the `cig_tax12_3` variable in the `states` data frame.

```{r eval=FALSE}
table(states$cig_tax12_3)
```

We can see that the names for these categories are LoTax, MidTax and HiTax. With the code below we use `str_replace_all()` to replace the characters with new characters, e.g. HiTax becomes High taxes.

```{r}
states$cig_taxes <- str_replace_all(states$cig_tax12_3, 
                                    c("HiTax" = "High taxes", 
                                      "MidTax" = "Middle taxes",
                                      "LoTax" = "Low taxes"))

table(states$cig_taxes)
```

For examples on more of the functions available in the `stringr` package, see [this introduction](https://cran.r-project.org/web/packages/stringr/vignettes/stringr.html).

## Factors

For the cigarette taxes we have worked with above, these are categorical data that we can order. To work with ordered and unordered categories, factors is a class in `R` class that makes these categories good to work with. For factors, we are going to use the package `forcats`. This package is also part of the `tidyverse`.

```{r}
library("forcats")
```

We create a new variable, `cig_taxes_cat` as a factor variable and then we see what levels we have (and the order of these).

```{r}
states$cig_taxes_cat <- factor(states$cig_taxes)

levels(states$cig_taxes_cat)
```

As we can see, these levels are now in the wrong order (sorted alphabetically). We can use the `fct_relevel()` to specify the order of the categories (from low to high).

```{r}
states$cig_taxes_cat <- fct_relevel(states$cig_taxes_cat, 
                                    "Low taxes", 
                                    "Middle taxes", 
                                    "High taxes")

levels(states$cig_taxes_cat)
```

This will become useful later on when we want to make sure that the categories in a data visualisation has the correct order.

For additional guidance on the functions available in the `forcats` package, see [https://forcats.tidyverse.org/](https://forcats.tidyverse.org/).

## Dates and time

To work with dates and time in `R`, there are two useful packages. The first is `hms` that is good with **h**ours, **m**inutes and **s**econds. The second is `lubridate` that is good with dates. Let us take a closer look at how to work with seconds, minutes and hours by loading the package `hms`.

```{r}
library("hms")
```

This package is useful if you want to easily convert minutes into hours, for example 500 minutes into hours. As `lubridate` also have an `hms()`, we will use `::` to tell that we are using the function `hms()` in the `hms` package.

```{r}
hms::hms(min = 500)
```

As we can see, this gives 8 hours and 20 minutes. We can also specify hours, minutes and seconds to get the time as POSIXct.

```{r}
hms::hms(hour = 15, min = 90, seconds = 12)
```

For dates, we will first load the package `lubridate`.

```{r}
library("lubridate")
```

This package has several functions that are useful in terms of working with dates. For example, we can use `ymd()` if we have text that has the year, month and day.

```{r}
ymd("2019/09/30")
```

The package can also work with months as text, such as:

```{r}
mdy("September 30, 2019")
```

The good thing about this is that we can work with the date information. Let us first save the date in an object called `date` and use `year()` to get the year out of the variable.

```{r}
# Save September 30, 2019 in object
date <- ymd("2019-09-30")

# Get year
year(date)
```

Similarly, we can get the week number out of the date.

```{r}
week(date)
```

We can see that this date was in week 39. We can use `wday()` to get the number of the day in the week this was.

```{r}
wday(date)
```

If we would rather prefer the name of the day, we can use `label` and `abbr` as options (the latter option in order to get the full day name).

```{r}
wday(date, label = TRUE, abbr = FALSE)
```

Some of the relevant functions in the `lubridate` package are: `year()` (year), `month()` (month), `week()` (week number), `day()` (day of month), `wday()` (day week), `qday()` (day of quarter), and `yday()` (day of year).
