---
output: html_document
editor_options: 
  chunk_output_type: console
---

# (PART) Presentation {-}

# Writing {#writing}

Good quantitative analysis is not only about doing stuff in `R`. In fact, you will not be showing your audience your results directly in `R`, but export the out in form of tables, figures, numbers, equations etc. Here, we will provide some advice on how to report your analysis.

## Equations

To transform statistical models into equations, i.e. to ensure that you can communicate the equation of the model you are estimating, we recommend that you use the `equatiomatic` package [@andersonetal2019]. You can read more about the package on [https://github.com/datalorax/equatiomatic](https://github.com/datalorax/equatiomatic). First, load the package (and install it if you haven't already done so).

```{r, message=FALSE, warning=FALSE}
library("equatiomatic")
```

To illustrate how the package works, let us estimate a multivariate OLS regression model with `mpg` as the outcome and `cyl` and `disp` as our predictors and save it in the object `mod1`.

```{r, message=FALSE, warning=FALSE}
mod1 <- lm(mpg ~ cyl + disp, data = mtcars)
```

To get the equation for this model, simply use `extract_eq()` on the object (in this case `mod1`).

```{r, message=FALSE, warning=FALSE}
extract_eq(mod1)
```

We can then take this LaTeX equation and put it into our document and get the equation (the built-in equation editor in Word also works with LaTeX equations):

$$
\text{mpg} = \alpha + \beta_{1}(\text{cyl}) + \beta_{2}(\text{disp}) + \epsilon
$$

If we would like to report the actual coefficients rather than alpha and betas, we can set the option `use_coefs` to `TRUE`.

```{r, message=FALSE, warning=FALSE}
extract_eq(mod1, use_coefs = TRUE)
```

The function works with all models supported by `tidy()` in the `broom` package. Noteworthy, it also works well with interaction terms. To show this, we can specify an interaction between `cyl` and `disp` in a linear model saved in the object `mod2`.

```{r, message=FALSE, warning=FALSE}
mod2 <- lm(mpg ~ cyl*disp, mtcars)
```

Again, we can use `extract_eq()` to get the equation for the model:

```{r, message=FALSE, warning=FALSE}
extract_eq(mod2)
```

This LaTeX equation can, as pointed out above, simply be placed into our document:

$$
\text{mpg} = \alpha + \beta_{1}(\text{cyl}) + \beta_{2}(\text{disp}) + \beta_{3}(\text{cyl} \times \text{disp}) + \epsilon
$$

# Tables

## Regression tables

To export regression tables from `R`, we are going to use the package `stargazer` [@hlavac2015]. Remember to install the package if you haven't already done so.

```{r, message=FALSE, warning=FALSE} 
library("stargazer")
```

First, we use the `stargazer()` function to show the output from the object `reg_obama` estimated in the OLS regression chapter. Notice that we also add the option `type = "text"`. If we do not do that, we will get the output as LaTeX code. 

```{r, message=FALSE, warning=FALSE} 
stargazer(reg_obama, type = "text")
```

This shows the output from one regression model. To add more regression models to the table, simply add a comma and the name of the object with the model. Below we use the same code as above and add the model with control variables included, `reg_obama_full`.

```{r, message=FALSE, warning=FALSE} 
stargazer(reg_obama, reg_obama_full, type = "text")
```

### Exporting the regression table

To export the regression table, we use the option `out` to specify, where we want to save our regresion table. Below we save the table in the file `tab-regression.htm`.

```{r, eval=FALSE, message=FALSE, warning=FALSE}
stargazer(reg_obama, reg_obama_full, 
          type = "text",
          out="tab-regression.htm")
```

An `.htm` file is a HTML file you can open in your browser (e.g. Google Chrome). To get it into Word, simply open the file via Word. You might have to do some extra changes before it is ready for a broader audience. Always try to make your tables look like tables in published articles and books.
